---
title: "p8105_hw3_av2718"
author: "Angelica Vina Albarracin"
date: "2022-10-15"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, message = FALSE)
```

```{r load_libraries}
library(tidyverse)   #load packages 
library(dplyr)
```

## Problem 2

This problem focuses on the Accelerometers data. Below we import and clean data from `accel_data.csv`. The process begins with importing the data and replacing each space in the original variable names with `snake_case`:

```{r}
library(readr)
accel_data = read_csv("data/accel_data.csv") %% #import data
  janitor::clean_names() %>% # tidy variable names
  mutate_if(is.double, as.numeric) #if double mutate variable to numeric

head(accel_data, 8) #view first 8 rows of Accelorometers data
```

The following code chuck we add a new a weekday vs weekend variable; and encode data with reasonable variable classes. The urine sample ID is spread across three columns, which correspond to three observation times. The original column names were informative but we don’t need to keep for the statistical analysis. To organize the data, we will remove some of the columns and then fix this issue using pivot_longer:

# ```{r}
# accel_data_tidy = 
#   pivot_longer(
#     accel_data,
#     activity_1:activity_1440,
#     names_to = "activity",
#     values_to = "activity_data",
#     names_prefix = "activity_"
#   ) %>% 
#   mutate(.data = ., 
#          week_type = if_else(day %in% c("Saturday", "Sunday"), "weekend", "weekday"),
#          day = fct_relevel(day, "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")) 
# 
# ```

