---
title: "p8105_hw3_av2718"
author: "Angelica Vina Albarracin"
date: "2022-10-15"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, message = FALSE)
```

```{r load_libraries}
library(tidyverse)   #load packages 
library(dplyr)
```

## Problem 2

This problem focuses on the Accelerometers data. Below we import and clean data from `accel_data.csv`. The process begins with importing the data and cleaning the variables names:

```{r}
library(readr)
accel_data = read_csv("data/accel_data.csv") %>% #import data
  janitor::clean_names() %>% # tidy variable names
  mutate_if(is.double, as.numeric) #if double mutate variable to   numeric
```

We quickly recognize a problem: the `activity` data is spread across 1443 columns, which correspond to 1443 activity counts for each minute of a 24-hour day. To address this and organize the data, we will use `pivot_longer`to wangle the activity data and create two new variables `workweek` and `weekday`:


```{r}
accel_data_tidy =  
  pivot_longer(   # organize activity data 
    accel_data,
    activity_1:activity_1440,
    names_to = "time_course",
    values_to = "activity_count",
    names_prefix = "activity_"                  
  ) %>%
  mutate(.data = .,  # new weekday vs weekend variable
         workweek = if_else(day %in% c("Saturday", "Sunday"), "weekend", "weekday"),
         weekday = fct_relevel(day, "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")) 

head(accel_data_tidy, 8) #view first 8 rows of tidied data
```

The resulting dataset has only 6 variables: `r ls(accel_data_tidy)`, and a size of `r dim(accel_data_tidy)`. Using this tidied dataset, we will now aggregate across minutes to create a `total_activity` variable for each day, and create a table showing these totals:

```{r}
total_activity = accel_data_tidy %>%  
  group_by(day,week) %>%           #Group total activity by day and week 
  summarize(day_total = sum(activity_count)) %>% 
  mutate(week = as.character(week))


pivot_wider(                 #Create friendly table for "human readers"
  total_activity,
  names_from = "day",
  values_from = "day_total"
)

```

As expected, from the table, we can observe different total activity levels depending on the day of the week. The patient seems to be more active during the weekdays. However, the data presented in this table isn't enough to make any conclusions. We would need to conduct further analysis. Nonetheless, from this table, we can observe on Saturday, in weeks 4 and 5, two unusual measurements, possibly due to a device malfunction. It's good that we noticed this issue, as it would affect our analysis and results.  

We will now use what we have learned in class to the inspection activity over the course of the day. In the code chuck below we will create a single-panel `plot` that shows the 24-hour activity time courses for each day and use color to indicate the day of the week:


```{r}
f1 = ggplot(accel_data_tidy, aes(x = time_course, y = activity_count, color = day))+
  geom_line()+
  theme_minimal()+
  theme(legend.position = "bottom")+  # Legend in bottom
  theme(plot.title = element_text(hjust = 0.5))+ # Center ggplot title
  labs(
    title = "24-hour Accelerometer Activity Count for Each Week Day",
    x = "Time Course (Minutes)",
    y = "Activity Count",
  )

f1
```

Based on this graph, we can say that the patient's physical activity fluctuates across the day. In general, we observe peak activity early in the day and a decrease in activity count during the middle of the day. During the weekdays, the patient seems to be more active in the morning and less active at night. On Saturday and Sunday, we observe more physical activity both in the morning and at night. However, it's difficult to describe a specific pattern or make a definite conclusion based on this graph alone. 

